---
output: 
  word_document:
    toc: true
    toc_depth: 3
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Compressed, Large-Language-Model Embedded Datasets of ICD-10-CM Descriptions

## Citing this work

(add link to bioaRxiv)

## License 

The code in this repository is licensed under [GPL v2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html) and the data
are licenced under [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/).

# Funding

This work was supported by the National Institute on Aging of the National Institutes of Health (NIH) through a project grant to Yale University (1R01AG071528). The organizations funding this study had no role in the design or conduct of the study; in the collection, management, analysis, or interpretation of the data; or in the preparation, review, or approval of the manuscript. The content of this publication is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health, the Department of Veterans Affairs, or the United States government. 

## ICD-10-CM Datasets

### 2022

1. [ICD-10-CM, 10-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2022-0010.csv?raw=true)
1. [ICD-10-CM, 50-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2022-0050.csv?raw=true)
1. [ICD-10-CM, 100-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2022-0100.csv?raw=true)
1. [ICD-10-CM, 1000-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2022-1000.csv?raw=true)

### 2021

1. [ICD-10-CM, 10-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2021-0010.csv?raw=true)
1. [ICD-10-CM, 50-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2021-0050.csv?raw=true)
1. [ICD-10-CM, 100-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2021-0100.csv?raw=true)
1. [ICD-10-CM, 1000-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2021-1000.csv?raw=true)

### 2020

1. [ICD-10-CM, 10-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2020-0010.csv?raw=true)
1. [ICD-10-CM, 50-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2020-0050.csv?raw=true)
1. [ICD-10-CM, 100-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2020-0100.csv?raw=true)
1. [ICD-10-CM, 1000-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2020-1000.csv?raw=true)

### 2019

1. [ICD-10-CM, 10-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2019-0010.csv?raw=true)
1. [ICD-10-CM, 50-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2019-0050.csv?raw=true)
1. [ICD-10-CM, 100-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2019-0100.csv?raw=true)
1. [ICD-10-CM, 1000-dimensions](https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/embedding-data/icd-10-cm-2019-1000.csv?raw=true)

## Background and introduction

The ICD-10-CM (International Classification of Diseases, 10th Revision, Clinical Modification) is a standardized classification system for diagnosing diseases, disorders, and health conditions. ICD-10 was developed by the World Health Organization (WHO) and adapted for use in the United States as [ICD-10-CM](https://www.cdc.gov/nchs/icd/icd-10-cm.htm) by the National Center for Health Statistics (NCHS). The standard plays a crucial role in the analysis of electronic medical records (EMRs) or electronic health records (EHRs) for several reasons:

1. Consistency and Standardization: The ICD-10-CM allows for a consistent and standardized method of coding and documenting medical conditions across healthcare providers and facilities. This helps to ensure accurate and uniform data exchange, analysis, and comparison.
2. Data Analysis and Research: The ICD-10-CM codes can be used to analyze patient data for clinical research, epidemiological studies, and public health surveillance. It helps to identify trends and patterns in diseases, monitor the effectiveness of treatments, and develop better prevention and management strategies.
3. Quality Measurement and Improvement: ICD-10-CM codes can be used to evaluate the quality of care provided by healthcare facilities, monitor patient outcomes, and identify areas for improvement. This information can be used to enhance the overall healthcare delivery system.
4. Reimbursement and Billing: ICD-10-CM codes play a vital role in healthcare reimbursement by providing a standardized method to classify and report medical conditions. Insurance companies and other payers use these codes to determine appropriate payments for medical services rendered.
5. Health Policy and Planning: ICD-10-CM codes help health authorities and policymakers to identify population health needs, allocate resources, and develop targeted healthcare policies and interventions.

While ICD-10-CM codes do provide a consistent and comprehensive set of categories, their incorporation into statistical and machine learning analyses can be challenging for several reasons. First, in the 2019 version of the standard there are 71,932 categories and that number has increased since then with the 2022 version containing 72,750 categories. As a result, analyses using these codes, where the set of codes is not restricted to smaller set, must take into account their high dimensionality or will require a large number of training samples in order to fit consistent models. Second, categorical variables are usually incorporated into analyses with a contrast encoding such as treatment, one-hot, helmert, etc. Contrast numeric representations are orthogonal or, under appropriate statistical assumptions, independent. However, ICD-10-CM codes represent a hierarchical structure, where codes are organized into chapters, blocks, and categories based on the type and anatomical location of the diseases or conditions. Applying traditional contrast encoding methods like (eg. one-hot, treatment, etc.) may not fully capture this hierarchical information, potentially resulting in a loss of valuable context and relationships between codes.

Researchers have considered alternative encoding methods or feature extraction techniques that can better represent the hierarchical structure of ICD-10-CM codes. However, incorporating both hierarchical structure and other contextual information in a general way can be difficult. The previous generation of word embeddings, which provide vector-encodings of words were shown effective for these types of tasks, with models like [med2vec providing improved abilities to predict patient mortality](https://academic.oup.com/jamiaopen/article/4/1/ooab022/6172949?login=false). Despite their advantages, word embeddings also have certain limitations. First, word embeddings are typically generated at the word or code level, which may not always be sufficient for capturing the hierarchical structure of ICD-10-CM codes. Second, the quality and representativeness of the word embeddings depend on the training data used to generate them. If the training data does not adequately cover the entire spectrum of medical conditions or encounters, the embeddings may not capture all relevant relationships or information. Third, many word embeddings do not account for the specific context in which a word appears. This can limit their ability to capture the nuances of medical conditions and the relationships between them.

Large language models (LLMs), address some of the shortcomings of traditional word embeddings through a combination of advanced techniques and architectures. Unlike traditional word embeddings that generate static representations, large language models generate contextualized embeddings. These embeddings take into account the surrounding words or tokens, allowing for a more nuanced representation of words and codes in different contexts. This helps in capturing the semantic relationships between codes more effectively. These models are pre-trained on vast amounts of text data, allowing them to learn general language representations before being fine-tuned for specific tasks. This pre-training enables the models to leverage existing knowledge and adapt more effectively to new tasks, even with limited task-specific data. Large language models can be incrementally updated or fine-tuned with new data, allowing them to adapt to evolving medical knowledge and practices more effectively than static word embeddings. And, while not explicitly designed for hierarchical data like ICD-10-CM codes, large language models can implicitly learn hierarchical relationships through their deep architectures and the context in which codes appear. This can help capture different levels of granularity and relationships between codes more effectively than traditional word embeddings.

This paper describes data sets provided as csv files, mapping ICD-10-CM codes to embeddings (a numeric vector of values), based on their descriptions. The embeddings were generated using the [BioGPT Large Language Model](https://academic.oup.com/bib/article/23/6/bbac409/6713511?guestAccessKey=a66d9b5d-4f83-4017-bb52-405815c907b9&login=false), which was trained on the biomedical literature including PubMed, PubMed Central, and clinical notes from MIMIC-III. This model was shown to do a better job of encoding context and relational information than competitors in the medical domain. Since the dimension of the embedding LLM is of high dimension (42,384), we provide dimension-reduced versions in 1,000, 100, 50, and 10 dimensions. The model generating the data were validated in two ways. The first way validates the dimension reduction. The embedding data were compressed using an auto-encoder. The out-of-sample accuracy of a validation set is examined as well as the performance of the model for other versions (by year) of the ICD-10-CM specification. Our results show that we can reduce the dimension of the data down to as few as 10 dimensions while maintaining the ability to reproduce the original embeddings, with the fidelity decreasing as the reduced-dimension representation decreases. The second way validates the conceptual representation by creating a supervised model to estimate the ICD-10-CM hiearchical categories. Again we see as the dimension of the compressed representation decreases the model accuracy decreases. Since multiple compression levels are provided, users are free to choose whichever suits their needs, allowing them to trade off accuracy for dimensionality.

The paper proceeds as follows. The next section provides a high-level description of the BioGPT and the embedding along with the construction of the autoencoder used to reduce the dimension of the embedding representation. That section then provides validation for both the dimension reduction as well as the representation. The third section provides an example of how to use the dataset to cluster ICD-10-CM codes using the R programming environment. The final section describes how to reproduce all of the results presented here.

The data sets and code use to generate them are available at https://github.com/kaneplusplus/icd-10-cs-embedding. The data are licensed under the [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/legalcode). The code is 
licensed under [GPL-v2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)

## Model description and performance

The provided data are generated by embedding ICD-10-CM descriptions using the BioGPT-Large model, which comprises 1.5 billion parameters and is accessible via the Hugging Face repository. The embedding process involves tokenizing textual phrases into tokens (words, subwords, or characters) and mapping them to unique vocabulary IDs. Token IDs are passed through an embedding layer, resulting in a sequence of continuous embedding vectors. Positional encodings are added element-wise to these vectors, enabling the model to capture token order and relative positions. The embeddings are then contextualized by passing them through the model's layers. An attention mask selectively controls information flow in the attention mechanism, allowing the model to weigh the importance of input tokens when generating contextualized embeddings in a 42,384 dimensional space.

The embedding is then compressed using an auto encoder. The auto-encoder is a series of
fully connected layers where the number of hidden nodes is approximately one order of magnitude
smaller than the previous layer and then an order of magnitude larger until the output layer.
For example, the autoencoder compressing to 10 dimensions has layers of size 42,384, 1,000,
100, 50, 10, 50, 100, 1,000, 42,384. Models whose dimension is large use them same
structure while retaining only the appropriate layers.

### Validating the dimension reduction

```{r model_perf, message = FALSE, warning = FALSE, echo = FALSE, fig.cap = "The autoencoder performance diagnostics ordered by decreasing Validation Loss."}
library(dplyr)
x = readRDS("model-performance.rds") 
x = x |>
  mutate(best_valid_loss = round(best_valid_loss, 3),
         best_train_loss = round(best_train_loss, 3)) |>
  select(embedding_dim, batch_size, best_train_loss, best_valid_loss)
names(x) = c("Embedding Dimension", "Batch Size","Training Loss", "Validation Loss")
knitr::kable(x, caption = "The autoencoder parameters and performance ordered by increasing validation loss.")
```

The autoencoder compressing the LLM embedding was fit on the 2019 ICD-10-CM descriptions
for 20 epochs, with batch sizes 64, 128, and 256, mean-square error loss between
the embedding and autoencoder estimate, and a validation data set comprised
of random subset of 10\% of the samples. The model performance is shown above.
Based on these results the models with the best validation loss where selected for distribution.

```{r autoencoder_perf, echo = FALSE, fig.cap = "The autoencoder year-validation diagnostics ordered by year."}
x = readRDS("year-validation.rds")
x = x |>
  mutate(pred_error = round(pred_error, 3),
         cod = round(cod, 3))
names(x) = c("Year", "Embedding Dimension", "MSE", "Coef. of Determination")
knitr::kable(x, caption = "The autoencoder year validation performance ordered by year.")
```

In addition to the 2019 validation the models selected for distribution were
tested on the 2020-2022 data sets to ensure their performance is comparable
over years. It should be noted that the ICD-10-CM codes do not vary much from
one year to the the next,
so we should not expect large differences. As expected, the mean square error
and coefficients of determination are similar to the 2019 data.

### Validating the embedding representation

As a final step in the validation process, we use the fact that in addition to
the description, the ICD-10-CM codes themselves carry hierarchical information,
which can be used to ensure that conceptual relationships are preserved
in the compressed embeddings. In particular, the leading letter and two 
numeric values categorize codes. For example, codes A00-B99 correspond to
infectious and parasitic diseases, C00-D49 correspond to neoplasms, etc. We
can therefore ensure that at least some of the relevant relationships are 
preserved in the compressed embedding representation by confirming that
the categories can be estimated at a rate higher than chance using a 
supervised model. Furthermore, we can quantify how much relevant predictive
information is lost in lower-dimensional representations.

The training data consists of a one-hot encoding of the ICD-10-CM
categories as the dependent variable and the compressed embedding values as
the indepedent values. The model consisted of two hidden layers with 100
nodes each. The loss function selected was categorical cross-entropy. The
model was trained using 30 epochs and a validation data set comprised of 10% 
samples, chosen at random. The performance in terms of both the accuracy and
the balanced accuracy is shown in the table below. As
with most problems of this type, compression of the data corresponds to
an increase in lost predictive information. 

```{r sm_perf, echo = FALSE, fig.cap = "The supervised model performance."}
x = readRDS("sup-model-perf.rds")
x = x |>
  mutate(accuracy = round(accuracy, 3),
         bal_accuracy = round(bal_accuracy, 3),
         emb_dim = as.integer(emb_dim))
names(x) = c("Embedding Dimension", "Accuracy", "Balanced Accuracy")
knitr::kable(x, caption = "The supervised models' performance ordered by increasing embedding dimension.")
```

Of note, the goal in presenting these results is not to necessarily to 
maximize the prediction accuracy. Rather, it is to show that the embedding retains the
hierarchical information in the ICD-10-CM codes. Some of the codes correspond to
conditions that could be classified in several ways, and as a result coding
for at least some of the conditions might be considered non-systematic.
Based on this criterion, we can conclude the embedding does retain much of the 
structural and conceptual information denoted in the descriptions, at least in 
terms of mapping to key categories of diseases and conditions.

## An example using the embedding data in R

To conclude, we present a simple example of how one might use the embedding
information in the R programming environment. Suppose we would like to 
visualize the ICD-10-CM codes beginning with G (diseases of the nervous system), 
I (diseases of the circulatory system), J (diseases of the respiratory system), 
and K (diseases of the digestive system) to better understand the relationships
between these categories or specific conditions in the the 50-dimensional 
embedding. For convenience, the projects page includes an `.rds` file
containing the available embeddings along with their URLs, which can be 
retrieved from the R console. The code categores can then be visualized 
by performing another dimension reduction (in this case we will use the
Rtsne package), to 2 dimensions and presented them
to a scatter plot as shown below. 


```{r message = FALSE, eval = TRUE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(Rtsne)
library(stringr)

# Download the locations of the embeddings.
dl = readRDS(
#  "https://github.com/kaneplusplus/icd-10-cm-embedding/blob/main/icd10_dl.rds"
  "icd10_dl.rds"
)

dl$url[dl$year == 2019 & dl$emb_dim == 50] = "embedding-data/icd-10-cm-2019-0050.csv"

# Read in the unspecified injury codes.
icd10s = read_csv(dl$url[dl$year == 2019 & dl$emb_dim == 50]) |>
  filter(str_detect(code, "^(G|I|J|K)")) |>
  mutate(desc = tolower(desc)) |>
#  filter(str_detect(desc, "^unspecified")) |>
  mutate(`Leading Letter` = str_sub(code, 1, 1)) |>
  distinct(pick(starts_with("V")), .keep_all = TRUE)

# Fit tSNE to the embedding.
tsne_fit = icd10s |> 
  select(starts_with("V")) |>
  scale() |>
  Rtsne(perplexity = 10)

# Bind the tSNE values to the data set.
icd10p = bind_cols(
  icd10s |>
    select(-starts_with("V")),
  tsne_fit$Y |>
    as.data.frame() |>
    rename(tSNE1="V1", tSNE2="V2") |>
    as_tibble()
)

# Visualize the results.
ggplot(icd10p, aes(x = tSNE1, y = tSNE2, color = `Leading Letter`)) +
  geom_point() +
  theme_minimal()
```

The visualization shows that a subset of the circulatory diseases (I) and
nervous system diseases (G) are well-differentiated from other conditions. It
also shows overlap between other conditions related to K (digestive diseases), 
J (respiratory diseases), and I (circulatory).

## Reproducing these results

R version: >= 4.2

R package dependencies:

  - `torch`
  - `reticulate`
  - `dplyr`
  - `tidyr`
  - `purrr`
  - `foreach`
  - `itertoos`
  - `readr`
  - `luz`
  - `tidyr`
  - `tibble`
  - `progress`
  - `stringr`
  - `yardstick`
  
Scripts

- `0-make-embeddings.R`
  - Purpose - create the embeddings created by BioGPT-Large
  - Dependencies
    - A conda evironment with the `torch` and `transformers` packages (see the `make-biogpt-conda-env` script)
  - Inputs
    - `icd-10-cm-codes/icd10cm_codes_2019.txt`
    - `icd-10-cm-codes/icd10cm_codes_2020.txt`
    - `icd-10-cm-codes/icd10cm_codes_2021.txt`
    - `icd-10-cm-codes/icd10cm_codes_2022.txt`
  - Outputs
    - An `icd-10-cm-embeddings` directory with subdirectories corresponding to each year, and subsubdirectories with files whose names correspond to the ICD-10-CM code holding R .rds files with the code, description, and BioGPT embedding values stored as a `data.frame`.
- `1-compress-icd-10-embeddings.R`
  - Purpose - recreate the embeddings created by BioGPT
  - Dependencies
    - R files: `autoencoder.R`
  - Inputs
  - Outputs
- `2-validation.R`
  - Purpose - recreate the embeddings created by BioGPT-Large
  - Dependencies
    - R files: `autoencoder.R`
  - Inputs
    - Files in the `icd-10-cm-embeddings/2019` directory.
  - Outputs
    - `model-performance.rds` holding a `data.frame` consisting of the model performance table.
    - Files in the `autoencoder-models` directory containing model to create the compressed embeddings.
- `3-create-datasets.R`
  - Purpose - recreate the embeddings created by BioGPT-Large
  - Dependencies
    - R files: `autoencoder.R`
  - Inputs
    - Files in the `autoencoder-models` directory.
    - Files in the `icd-10-cm-embeddings` directory for all years (2019-2020).
  - Outputs
    - `year-validation.rds` holding a data frame of the autoencoder year-validation model performance.
    - Files in the `embedding-data` directory holding the embedding values as .csv files for all year-dimension combinations.
- `4-estimate-leading-char.R`
  - Purpose - recreate the embeddings created by BioGPT-Large
  - Dependencies
    - R files: `alpha-char-model.R`
  - Inputs
    - Files in the `embedding-data` directory.
  - Outputs
    - Files in the `luz-supervised-models` directory holding the `luz` package representation of the fitted models.
    - The `supervised-model-perf.rds` files containing a `data.frame` summarizing the supervised model performance.

© Michael J. Kane (kaneplusplus at proton mail dot com)
